---
title: "nyc-taxi-duration-eda-prediction"
author: "Doron Dusheiko"
date: "20 September 2018"
output: html_document
number_sections: true 
toc: true 
fig_width: 7 
fig_height: 4.5 
theme: cosmo 
highlight: tango 
code_folding: show 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a comprehensive Exploratory Data Analysis for the New York City Taxi Trip Duration competition with tidy R and ggplot2.

The goal of this playground challenge is to predict the duration of taxi rides in NYC based on features like trip coordinates or pickup date and time. The data comes in the shape of 1.5 million training observations (../input/train.csv) and 630k test observation (../input/test.csv). Each row contains one taxi trip.

In this notebook, we will first study and visualise the original data, engineer new features, and examine potential outliers. Then we add two external data sets on the NYC weather and on the theoretically fastest routes. We visualise and analyse the new features within these data sets and their impact on the target trip_duration values. Finally, we will make a brief excursion into viewing this challenge as a classification problem and finish this notebook with a simple XGBoost model that provides a basic prediction.

Enough talk. Let’s get started:

## Load libraries and helper functions

```{r}
library('ggplot2') # visualisation
library('scales') # visualisation
library('grid') # visualisation
library('RColorBrewer') # visualisation
library('corrplot') # visualisation
library('alluvial') # visualisation
library('dplyr') # data manipulation
library('readr') # input/output
library('data.table') # data manipulation
library('tibble') # data wrangling
library('tidyr') # data wrangling
library('stringr') # string manipulation
library('forcats') # factor manipulation
library('lubridate') # date and time
library('geosphere') # geospatial locations
library('leaflet') # maps
library('leaflet.extras') # maps
library('maps') # maps
library('xgboost') # modelling
library('caret') # modelling
```

We use the multiplot function, courtesy of [R Cookbooks](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/) to create multi-panel plots.

```{r}
# Define multiple plot function

# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.

# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {

  # Make a list from the ... arguments and plotlist

  plots <- c(list(...), plotlist)
  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout

  if (is.null(layout)) {

    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols

    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))

  }
  
  if (numPlots==1) {

    print(plots[[1]])

  } else {

    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {

      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))

    }

  }

}
```

## Load data

We use _data.table’s_ fread function to speed up reading in the data:

```{r}
train <- as.tibble(fread('input/train.csv'))
test <- as.tibble(fread('input/test.csv'))
sample_submit <- as.tibble(fread('input/sample_submission.csv'))
```

## File structure and content

Let’s have an overview of the data sets using the summary and glimpse tools. First the training data:

```{r}
summary(train)
```

```{r}
glimpse(train)
```

And then the testing data:

```{r}
summary(test)
```

```{r}
glimpse(test)
```

We find:

- `vendor_id` only takes the values 1 or 2, presumably to differentiate two taxi companies
- `pickup_datetime` and (in the training set) dropoff_datetime are combinations of date and time that we will have to re-format into a more useful shape
- `passenger_count` takes a median of 1 and a maximum of 9 in both data sets
- The `pickup/dropoff_longitute/latitute` describes the geographical coordinates where the meter was activate/deactivated.
- `store_and_fwd_flag` is a flag that indicates whether the trip data was sent immediately to the vendor (“N”) or held in the memory of the taxi because there was no connection to the server (“Y”). Maybe there could be a correlation with certain geographical areas with bad reception?
- `trip_duration`: our target feature in the training data is measured in seconds.

## Missing values

Knowing about missing values is important because they indicate how much we don’t know about our data. Making inferences based on just a few cases is often unwise. In addition, many modelling procedures break down when missing values are involved and the corresponding rows will either have to be removed completely or the values need to be estimated somehow.

Here, we are in the fortunate position that our data is complete and there are no missing values:

```{r}
sum(is.na(train))

sum(is.na(test))
```

## Combining train and test

In preparation for our eventual modelling analysis we combine the train and test data sets into a single one. I find it generally best not to examine the test data too closely, since this bears the risk of overfitting your analysis to this data. However, a few simple consistency checks between the two data sets can be of advantage.

```{r}
combine <- bind_rows(train %>% mutate(dset = "train"), 
                     test %>% mutate(dset = "test",
                                     dropoff_datetime = NA,
                                     trip_duration = NA))

combine <- combine %>% mutate(dset = factor(dset))
```

## Reformating features

For our following analysis, we will turn the data and time from characters into date objects. We also recode vendor_id as a factor. This makes it easier to visualise relationships that involve these features.

```{r}
train <- train %>%
  mutate(pickup_datetime = ymd_hms(pickup_datetime),
         dropoff_datetime = ymd_hms(dropoff_datetime),
         vendor_id = factor(vendor_id),
         passenger_count = factor(passenger_count))
```

## Consistency check

It is worth checking whether the `trip_durations` are consistent with the intervals between the pickup_datetime and dropoff_datetime. Presumably the former were directly computed from the latter, but you never know. Below, the check variable shows “TRUE” if the two intervals are not consistent:

```{r}
train %>%
  mutate(check = abs(int_length(interval(dropoff_datetime, pickup_datetime)) + trip_duration) > 0) %>%
  select(check, pickup_datetime, dropoff_datetime, trip_duration) %>%
  group_by(check) %>%  count()
```

And we find that everything fits perfectly.

# Individual feature visualisations

Visualisations of feature distributions and their relations are key to understanding a data set, and they often open up new lines of inquiry. I always recommend to examine the data from as many different perspectives as possible to notice even subtle trends and correlations.

In this section we will begin by having a look at the distributions of the individual data features.

We start with a map of NYC and overlay a managable number of pickup coordinates to get a general overview of the locations and distances in question. For this visualisation we use the leaflet package, which includes a variety of cool tools for interactive maps. In this map you can zoom and pan through the pickup locations

```{r}
set.seed(1234)
foo <- sample_n(train, 8e3)

leaflet(data = foo) %>% addProviderTiles("Esri.NatGeoWorldMap") %>%
  addCircleMarkers(~ pickup_longitude, ~pickup_latitude, radius = 1,
                   color = "blue", fillOpacity = 0.3)
```

In turns out that almost all of our trips were in fact taking place in Manhattan only. Another notable hot-spot is JFK airport towards the south-east of the city.

The map gives us an idea what some of the our distributions could look like. Let’s start with plotting the target feature `trip_duration`:

```{r}
train %>%
  ggplot(aes(trip_duration)) +
  geom_histogram(fill = "red", bins = 150) +
  scale_x_log10() +
  scale_y_sqrt()
```

Note the logarithmic x-axis and square-root y-axis.

We find:

- the majority of rides follow a rather smooth distribution that looks almost log-normal with a peak just short of 1000 seconds, i.e. about 17 minutes.
- There are several suspiciously short rides with less than 10 seconds duration.
- Additionally, there is a strange delta-shaped peak of trip_duration just before the 1e5 seconds mark and even a few way above it:

```{r}
train %>%
  arrange(desc(trip_duration)) %>%
  select(trip_duration, pickup_datetime, dropoff_datetime, everything()) %>%
  head(10)
```

Those records would correspond to 24-hour trips and beyond, with a maximum of almost 12 days. I know that rush hour can be bad, but those values are a little unbelievable.

Over the year, the distributions of `pickup_datetime` and `dropoff_datetime` look like this:

```{r}
p1 <- train %>%
  ggplot(aes(pickup_datetime)) +
  geom_histogram(fill = "red", bins = 120) +
  labs(x = "Pickup dates")

p2 <- train %>%
  ggplot(aes(dropoff_datetime)) +
  geom_histogram(fill = "blue", bins = 120) +
  labs(x = "Dropoff dates")

layout <- matrix(c(1,2), 2, 1, byrow=FALSE)
multiplot(p1, p2, layout=layout)
```
```{r}
p1 <- 1; p2 <- 1
```

Fairly homogeneous, covering half a year between January and July 2016. There is an interesting drop around late January early February:

```{r}
train %>%
  filter(pickup_datetime > ymd("2016-01-20") & pickup_datetime < ymd("2016-02-10")) %>%
  ggplot(aes(pickup_datetime)) +
  geom_histogram(fill = "red", bins = 120)
```

That’s winter in NYC, so maybe snow storms or other heavy weather? Events like this should be taken into account, maybe through some handy external data set?

In the plot above we can already see some daily and weekly modulations in the number of trips. Let’s investigate these variations together with the distributions of `passenger_count` and `vendor_id` by creating a multi-plot panel with different components:

```{r}
p1 <- train %>%
  group_by(passenger_count) %>%
  count() %>%
  ggplot(aes(passenger_count, n, fill = passenger_count)) +
  geom_col() +
  scale_y_sqrt() +
  theme(legend.position = "none")

p2 <- train %>%
  ggplot(aes(vendor_id, fill = vendor_id)) +
  geom_bar() +
  theme(legend.position = "none")

p3 <- train %>%
  ggplot(aes(store_and_fwd_flag)) +
  geom_bar() +
  theme(legend.position = "none") +
  scale_y_log10()

p4 <- train %>%
  mutate(wday = wday(pickup_datetime, label = TRUE)) %>%
  group_by(wday, vendor_id) %>%
  count() %>%
  ggplot(aes(wday, n, colour = vendor_id)) +
  geom_point(size = 4) +
  labs(x = "Day of the week", y = "Total number of pickups") +
  theme(legend.position = "none")

p5 <- train %>%
  mutate(hpick = hour(pickup_datetime)) %>%
  group_by(hpick, vendor_id) %>%
  count() %>%
  ggplot(aes(hpick, n, color = vendor_id)) +
  geom_point(size = 4) +
  labs(x = "Hour of the day", y = "Total number of pickups") +
  theme(legend.position = "none")


layout <- matrix(c(1,2,3,4,5,5), 3, 2, byrow=TRUE)
multiplot(p1, p2, p3, p4, p5, layout=layout)
```

```{r}
p1 <- 1; p2 <- 1; p3 <- 1; p4 <- 1; p5 <- 1
```

We find:

- There are a few trips with zero, or seven to nine passengers but they are a rare exception:

```{r}
train %>%
  group_by(passenger_count) %>%
  count()
```

- The vast majority of rides had only a single passenger, with two passengers being the (distant) second most popular option.
- Towards larger passenger numbers we are seeing a smooth decline through 3 to 4, until the larger crowds (and larger cars) give us another peak at 5 to 6 passengers.
- Vendor 2 has significantly more trips in this data set than vendor 1 (note the logarithmic y-axis). This is true for every day of the week.
- We find an interesting pattern with Monday being the quietest day and Friday very busy. This is the same for the two different vendors, with `vendor_id == 2` showing significantly higher trip numbers.
- As one would intuitively expect, there is a strong dip during the early morning hours. There we also see not much difference between the two vendors. We find another dip around 4pm and then the numbers increase towards the evening.
- The `store_and_fwd_flag` values, indicating whether the trip data was sent immediately to the vendor (“N”) or held in the memory of the taxi because there was no connection to the server (“Y”), show that there was almost no storing taking place (note again the logarithmic y-axis):

```{r}
train %>%
  group_by(store_and_fwd_flag) %>%
  count()
```

These numbers are equivalent to about half a percent of trips not being transmitted immediately.

The trip volume per hour of the day depends somewhat on the month and strongly on the day of the week:

```{r}
p1 <- train %>%
  mutate(hpick = hour(pickup_datetime),
         Month = factor(month(pickup_datetime, label = TRUE))) %>%
  group_by(hpick, Month) %>%
  count() %>%
  ggplot(aes(hpick, n, color = Month)) +
  geom_line(size = 1.5) +
  labs(x = "Hour of the day", y = "count")

p2 <- train %>%
  mutate(hpick = hour(pickup_datetime),
         wday = factor(wday(pickup_datetime, label = TRUE))) %>%
  group_by(hpick, wday) %>%
  count() %>%
  ggplot(aes(hpick, n, color = wday)) +
  geom_line(size = 1.5) +
  labs(x = "Hour of the day", y = "count")

layout <- matrix(c(1,2),2,1,byrow=FALSE)
multiplot(p1, p2, layout=layout)
```

```{r}
p1 <- 1; p2 <- 1
```

We find:

- January and June have fewer trips, whereas March and April are busier months. This tendency is observed for both `vendor_ids`.
- The weekend (Sat and Sun, plus Fri to an extend) have higher trip numbers during the early morning ours but lower ones in the morning between 5 and 10, which can most likely be attributed to the contrast between NYC business days and weekend night life. In addition, trip numbers drop on a Sunday evening/night.

Finally, we will look at a simple overview visualisation of the pickup/dropoff latitudes and longitudes:

```{r}
p1 <- train %>%
  filter(pickup_longitude > -74.05 & pickup_longitude < -73.7) %>%
  ggplot(aes(pickup_longitude)) +
  geom_histogram(fill = "red", bins = 40)

p2 <- train %>%
  filter(dropoff_longitude > -74.05 & dropoff_longitude < -73.7) %>%
  ggplot(aes(dropoff_longitude)) +
  geom_histogram(fill = "blue", bins = 40)

p3 <- train %>%
  filter(pickup_latitude > 40.6 & pickup_latitude < 40.9) %>%
  ggplot(aes(pickup_latitude)) +
  geom_histogram(fill = "red", bins = 40)

p4 <- train %>%
  filter(dropoff_latitude > 40.6 & dropoff_latitude < 40.9) %>%
  ggplot(aes(dropoff_latitude)) +
  geom_histogram(fill = "blue", bins = 40)

layout <- matrix(c(1,2,3,4),2,2,byrow=FALSE)
multiplot(p1, p2, p3, p4, layout=layout)
```

```{r}
p1 <- 1; p2 <- 1; p3 <- 1; p4 <- 1
```
Here we had constrain the range of latitude and longitude values, because there are a few cases which are way outside the NYC boundaries. The resulting distributions are consistent with the focus on Manhattan that we had already seen on the map. These are the most extreme values from the `pickup_latitude` feature:

```{r}
train %>%
  arrange(pickup_latitude) %>%
  select(pickup_latitude, pickup_longitude) %>%
  head(5)
```

```{r}
train %>%
  arrange(desc(pickup_latitude)) %>%
  select(pickup_latitude, pickup_longitude) %>%
  head(5)
```

We need to keep the existence of these (rather astonishing) values in mind so that they don’t bias our analysis.

# Feature relations

While the previous section looked primarily at the distributions of the individual features, here we will examine in more detail how those features are related to each other and to our target `trip_duration`.

## Pickup date/time vs trip_duration

How does the variation in trip numbers throughout the day and the week affect the average trip duration? Do quieter days and hours lead to faster trips? Here we include the vendor_id as an additional feature. Furthermore, for the hours of the day we add a smoothing layer to indicate the extent of the variation and its uncertainties:

```{r}
p1 <- train %>%
  mutate(wday = wday(pickup_datetime, label = TRUE)) %>%
  group_by(wday, vendor_id) %>%
  summarise(median_duration = median(trip_duration)/60) %>%
  ggplot(aes(wday, median_duration, color = vendor_id)) +
  geom_point(size = 4) +
  labs(x = "Day of the week", y = "Median trip duration [min]")

p2 <- train %>%
  mutate(hpick = hour(pickup_datetime)) %>%
  group_by(hpick, vendor_id) %>%
  summarise(median_duration = median(trip_duration)/60) %>%
  ggplot(aes(hpick, median_duration, color = vendor_id)) +
  geom_smooth(method = "loess", span = 1/2) +
  geom_point(size = 4) +
  labs(x = "Hour of the day", y = "Median trip duration [min]") +
  theme(legend.position = "none")


layout <- matrix(c(1,2),2,1,byrow=FALSE)
multiplot(p1, p2, layout=layout)
```

```{r}
p1 <- 1; p2 <- 1
```

We find:

- There is indeed a similar pattern as for the business of the day of the week. Vendor 2, the one with the more frequent trips, also has consistently higher trip durations than vendor 1. It will be worth adding the vendor_id feature to a model to test its predictive importance.
- Over the course of a typical day we find a peak in the early afternoon and dips around 5-6am and 8pm. The weekday and hour of a trip appear to be important features for predicting its duration and should be included in a successful model.

## Passenger count and Vendor vs trip_duration

The next question we are asking is whether different numbers of passengers and/or the different vendors are correlated with the duration of the trip. We choose to examine this issue using a series of boxplots for the passenger_counts together with a facet wrap which contrasts the two vendor_ids:

```{r}
train %>%

  ggplot(aes(passenger_count, trip_duration, color = passenger_count)) +
  geom_boxplot() +
  scale_y_log10() +
  theme(legend.position = "none") +
  facet_wrap(~ vendor_id) +
  labs(y = "Trip duration [s]", x = "Number of passengers")
```

We find:

- Both vendors have short trips without any passengers.
- Vendor 1 has all of the trips beyond 24 hours, whereas vendor 2 has all of the (five) trips with more than six passengers and many more trips that approach the 24-hour limit.
- Between 1 and 6 passengers the median trip durations are remarkably similar, in particular for vendor 2. There might be differences for vendor 1, but they are small (note the logarithmic y-axis):

```{r}
train %>%
  ggplot(aes(trip_duration, fill = vendor_id)) +
  geom_density(position = "stack") +
  scale_x_log10()
```

Comparing the densities of the trip_duration distribution for the two vendors we find that the medians are very similar, whereas the means are likely skewed by vendor 2 containing most of the long-duration outliers:

```{r}
train %>%
  group_by(vendor_id) %>%
  summarise(mean_duration = mean(trip_duration),
            median_duration = median(trip_duration))
```

## Store and Forward vs trip_duration

The temporary storing of the trip data only occured for Vendor 1:

```{r}
train %>%
  group_by(vendor_id, store_and_fwd_flag) %>%
  count()
```

```{r}
train %>%
  filter(vendor_id == 1) %>%
  ggplot(aes(passenger_count, trip_duration, color = passenger_count)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~ store_and_fwd_flag) +
  theme(legend.position = "none") +
  labs(y = "Trip duration [s]", x = "Number of passengers") +
  ggtitle("Store_and_fwd_flag impact")
```

We find that there is no overwhelming differences between the stored and non-stored trips. The stored ones might be slightly longer, though, and don’t include any of the suspiciously long trips.


